---
title: Exploring the {gratia} package
author: Eric R. Scott
date: '2021-02-04'
slug: [gratia-package]
categories: []
tags: [DLNMs, GAMs, R]
subtitle: ''
summary: ''
authors: []
lastmod: '2021-02-04T12:27:39-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
draft: yes
---

<!-- Hey, you might not bother finishing this.  You already talked about it in an earlier post on visualizing results of DLNMs -->

```{r include=FALSE, paged.print=FALSE}
library(tidyverse)
library(here)
ha <- read_rds(here("sitedata", "ha_sample.rds"))
```

{{% alert note %}}

This is part of series about distributed lag non-linear models.
Please read the [first post](/post/dlnm) for an introduction and a disclaimer.

{{% /alert %}}

In my last post in the DLNM series, I used the `gratia` package with no introduction or explanation.  I want to explore it a bit here.  In particular, I want to use it to explore the difference between plotting predicted values of a GAM and plotting a smooth itself.

## Plotting model results

There are several ways one could plot the effect of one covariate on the response for a given model.  One way is to plot the best-fit line between some covariate, $x$, and fitted values when all other covariates are set to their average (for continuous) or reference (for factors) levels.  I'd call this a "marginal effects plot" of the effect of $x$ on $y$, although I'm not 100% sure I'm correct with that definition.  We can also make one of these for a GAM using `predict()` or `broom::augment()`.  

```{r}
library(mgcv)
library(dlnm)
# same GAM as last time
growth <-
  gam(log_size_next ~ 
        s(log_size) +
        s(spei_history, L, #crossbasis function
          bs = "cb", 
          k = c(3, 24), 
          xt = list(bs = "cr")),
      family = gaussian(link = "identity"),
      method = "REML",
      data = ha)
```

I'll make a marginal effects plot for the effect of log(size) in year t on log(size) in year t+1.  First, I need to create `newdata`:

```{r paged.print=FALSE}
newdata <- 
  #create mesh of x values to predict over
  tibble(log_size = seq(min(ha$log_size), max(ha$log_size), length.out = 100)) %>% 
  #add covariate of SPEI at mean value
  mutate(spei_history = mean(ha$spei_history),
         L = matrix(1:36, nrow = 100, ncol = 36, byrow = TRUE))
```


Now I can use this to predict fitted values given average SPEI.

```{r paged.print=FALSE}
pred <- broom::augment(growth, newdata = newdata)
ggplot(pred, aes(x = log_size)) + 
  geom_line(aes(y = .fitted), color = "blue") + 
  geom_point(data = ha, aes(x = log_size, y = log_size_next), alpha = 0.5) +
  labs(x = "log(size_t)", y = "log(size_t+1)")
```

I can also create a marginal effects plot for the crossbasis smooth, but it is more complicated because I need to predict over all combinations of a range of SPEI values and lag times. This is what the `cb_margeff()` function I introduced in a [previous post]() does.

```{r}
newdata2 <- 
```



...

Then, in the next post, I'll use these plots to explore differences between DLNM's `bs = "cb"` and a standard tensor product smooth `te()`.


For a GAM like

y ~ 1 + s(size) + te(SPEI, L)

I could get the overall effect of lagged SPEI by just plotting `te(SPEI, L)` and adding the intercept and `s(mean(size))`.  So really I'd only need to do `predict()` on a single value of size as newdata.
